Lower Limits on the Core Mass for Giant Planets

### 2013-01-22 tex/draft/draft.tex
added math, tex and slides folders

% !TEX TS-program = pdflatexmk

@@

\usepackage{pdfsync}
\usepackage{subeqnarray}
\usepackage{natbib}

@@

\newcommand{\ie}{i.e.\ }
\newcommand{\eg}{e.g.\ }
\newcommand{\p}{\partial}
\newcommand{\xv}{\vc{x}}
\newcommand{\kv}{\vc{k}}
\newcommand{\brak}[1]{\langle #1\rangle}

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
%\newcommand{\au}{\; \varpi_{\rm AU}}

\newcommand{\AU}{\; {\rm AU}}
\newcommand{\yr}{\; {\rm yr}}
\def\K{\; {\rm K}}

@@

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

@@

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\co}{_{\rm c}}
\newcommand{\di}{_{\rm o}}
\newcommand{\cb}{_{\rm CB}}
\newcommand{\surf}{_M}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

@@

\shorttitle{Core Accretion}
\shortauthors{Youdin }

\title{Core Accretion of Self-Gravitating Atmospheres}
\author{Andrew N.\ Youdin}
\affil{Harvard Smithsonian Center for Astrophysics}

%\begin{abstract}
%\end{abstract}
\section{Introduction}
The core accretion model \citep{PerCam74, MizNak78} proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.    Early versions of the theory found a ``critical core mass", $\MC$, above which static solutions do not exist and unstable atmospheric collapse would occur \citep{Har78, Miz80}.  These works used hydrostatic models that include heating by planetesimal accretion, but neglect the heat generated by the gravitational, or Kelvin-Helmholtz (KH), contraction of the atmosphere.

In evolutionary models that include KH contraction \citep{BodPol86}, the atmosphere does not undergo a hydrodynamic collapse.  Instead the quasi-static contraction of the envelope accelerates continuously.  Contraction becomes particularly rapid after the the  ``crossover" mass, when core and atmospheric masses are equal \citep{pollack96}.  The concepts of critical and crossover masses are similar, and even agree numerically when the planetesimal accretion luminosity exceeds the KH luminosity.  However if the planetesimal accretion rate is small and KH contraction dominates, the critical core mass will be smaller than the more relevant  crossover mass \citep{IkoNak00}.

This paper will consider 1D atmospheres  where KH contraction dominates over planetesimal accretion.  Such solutions give realistic lower limits on the core mass required to accrete a massive gaseous atmosphere.  Indeed even models that include planetesimal accretion obtain the fastest formation times by stopping planetesimal accretion after the core reaches a certain mass \citep{HubBod05}.

The early phases of atmospheric growth are usually treated with 1D models where the outer boundary of the atmosphere matches onto the conditions in the disk midplane.   As the atmosphere continues to grow, 1D solutions are no longer adequate.  The atmosphere both notices that the disk is not spherically symmetric and at large enough masses will start to open a gap in the disk.  In this case the accretion of gas is limited by the ability of the disk to supply gas to the growing protoplanet \citep{DAnLub08}.  This later stage evolution is crucial for determining the final mass of the planet, but  not considered in our calculations of the initial collapse.

%Analytic models to explain the critical core mass have their shortcomings. \cite{Ste82} assumed a fully radiative atmosphere, which is inconsistent with realistic luminosities and opacities.  \cite{Wuc93} assumed a fully convective atmosphere which is is inconsistent with the ability of atmospheres to cool.  \citet[hereafter R06]{Raf06} considerably advanced the analytic understanding of core accretion models.  (Explain what's great.)  However R06 neglected the luminosity due to gas accretion.

%Most analyses of the critical core mass considers planetesimal accretion as the dominant luminosity source.  One notable exception is \cite{PapNel05}.  There are several reasons to consider the atmospheres on protoplanets that are not accreting solids.  First the planetesimal accretion rate may be a negligible source of luminosity.  (In this sense our results complement those of R06).  Second the growth of an atmosphere around a preexisting core represents lower limit to the timescale to form a giant planet.  Since atmospheric mass grows sharply with core mass, there may be little loss in assuming that atmospheric accretion begins once core formation stops.  Finally while it is desirable to self-consistently solve for core growth and atmospheric accretion, it is impossible to account for all possible core growth processes, which are quite complicated (refs).  
%For instance planet migration or scattering could bring a core that formed rapidly in one region of the disk to a region of the disk where cores could not form rapidly.

Section \ref{sec:structure} describes the structure of planet atmospheres embedded in a gas disk.   We then derive the critical core mass to trigger runaway collapse of the atmosphere in \S\ref{sec:critmass}.

\section{Disk Model}
As a fiducial disk model, we adopt the version of the minimum mass, passively irradiated model of \citet{cy10}, which gives the surface density and midplane temperature as
\begin{subeqnarray}
\label{eq_sigmag}
\varSigma  &=& 70 \,F_\varSigma \aun{10}^{-3/2} ~{\rm g~cm}^{-2} \\
T &=& 45  \,F_T\, \aun{10}^{-3/7} ~{\rm K}
\end{subeqnarray} 
where $\aun{10} \equiv a/{10~\rm AU}$ scales the disk radius and the normalization factors $F_\varSigma$ and $F_T$ adjust the disk mass and temperature relative to the canonical values.   The resulting disk midplane pressure 
\begin{equation}
P\di = 6.9 \times 10^{-9} F_\varSigma \sqrt{F_T m_\ast} \, \aun{10}^{-45/14}~{\rm bar}
\end{equation} 
for a mean molecular weight of 2.35.  The stellar mass in solar units is $m_\ast \equiv M_\ast/M_\odot$, which only includes the gravitational effects, though stellar mass also influences temperature and probably correlates with average disk mass.  Away from the midplane the pressure drops with height $z$ as a Gaussian with scale-height
\begin{equation}
H = {\sqrt{\Rg T\di} \over \varOmega} = 0.42 \sqrt{F_T \over m_\ast}  \, \aun{10}^{9/7} \AU\, .
\end{equation}

\section{Atmospheric Structure}\label{sec:structure}
In this section we review the structure of atmospheres surrounding a solid protoplanetary core embedded in a gas disk.  The core has mass $M\co$ and radius $R\co = [3 M\co/(4 \pi \rho\co)]^{1/3}$ with internal density $\rho\co \approx 2~{\rm g~cm^{-3}}$.  For protoplanetary disk parameters we use the minimum mass, passively irradiated disk model presented in \citet[hereafter  CY10]{cy10}.  The disk has a midplane gas density $\rho\di$ and temperature $T\di$.

The core can attract a dense atmosphere inside its Bondi radius
\begin{equation} 
R_B = {G M\co \over \mathcal{R} T\di} \approx 0.17 {\mcn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
\end{equation} 
where $\mcn{10} \equiv M\co/(10~M_\oplus)$ and $\mathcal{R}$ is the specific gas constant.

Low mass planets satisfy a hierarchy of length scales,  $R\co \ll R_B < R_{\rm H} < H$, where %$h\di \approx 3.3 \times 10^6 \au^{9/7}~{\rm km}$ is the gas scale-height and
\begin{equation} 
R_{\rm H} = \left(M\co \over 3 M_\ast \right)^{1/3}a \approx 0.22 {\mcn{10}^{1/3} \, \aun{10} \over m_\ast^{1/3}}~\AU
\end{equation} 
is the Hill radius, inside which the planet's gravity competes with Keplerian shear.

As protoplanets grow in mass the length scales $R_{\rm H} \sim R_{\rm B} \sim H$ are roughly equal at a characteristic ``transitional" mass
\begin{equation}
M_{\rm tr} > {(\Rg T\di)^{3/2} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
\end{equation} 
Technically the length scales re-order in three steps of increasing mass.  First for $M > M_{\rm tr}/\sqrt{3}$ the ordering becomes $R_{\rm H} < R_{\rm B} < H$ as tidal truncation becomes more important than thermal constriants.  While the shear flow past the atmosphere is clearly not spherically symmetric, it is still common take the atmosphere as spherically symmetric with an outer boundary at the Hill radius.  When the mass increases further to $M > M_{\rm tr}$ then  $R_{\rm H} < H < R_{\rm B}$, which does not change the fact that tidal truncation is the stronger effect at the outer boundary.

Finally when $M > 3 M_{\rm tr}$ the high mass scaling $H < R_{\rm H} < R_{\rm B}$ is attained.  The assumption of spherical symmetry fails for two reasons.  First the planet now notices the vertical density stratification of the disk.  Second, the planet's torques on the disk can begin to open gaps, creating further density inhomogeneity.  Since runaway accretion typically begins for masses well below $\sim 75 M_\oplus$, the assumption of spherical symmetry is a good, if not perfect, approximation for understanding whether cores can attrach massive atmospheres.  Moreover the approximation becomes even better in the disk's outer regions.

Another relevant for atmospheric structure is the mass of disk gas that fills the Bondi radius, at the ambient density.  The ratio of this mass to the core mass is 
\begin{equation}
\theta\co \equiv { 4 \pi \rho\di R_B^3 \over 3 M\co} \approx 5 \times 10^{-3} {\mcn{10}^2 \over \aun{10}^{3/2}}{F_\varSigma \sqrt{m_\ast} \over F_T^{7/2}}\, .
\end{equation} 
The smaller this parameter the more gravitational compression is required for an atmosphere to become self-gravitating.

There are two limiting cases of atmospheric structure, isothermal and adiabatic.  Atmospheres that are isothermal at $T\di$ have cooled completely and give the most massive atmosphere.   Convectively unstable atmospheres transport heat very efficiently and are nearly adiabatic.  Adiabatic atmospheres are the opposite extreme of the isothermal case, giving the steepest increase of temperature with atmospheric dept.  The least dense atmospheres are fully adiabatic with the entropy of the disk.

The structure of an embedded protoplanet consists of an inner adiabatic region that matches onto an outer radiative zone, that in some cases, can be approximated as nearly isothermal.  This basic structure was demonstrated by R06 for an atmosphere supported by planetesimal accretion luminosity.  PN05 showed that atmospheres powered by gas accretion alone have a similar structure.  Due to their intense irradiation, ``hot Jupiters" similarly have an isothermal layer that matches onto a deeper adiabatic interior \citep{ab06, ym10}.

\subsection{Isothermal Atmosphere}
We now consider the structure of a low mass, i.e.\ non-self gravitating, isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho\di$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$.  The resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho\di \exp \left({R_B \over r} - {1 \over n_B} \right) \approx   \rho\di \exp \left(R_B \over r  \right)
\end{equation} 
where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However the choice of boundary condition does have an order unity effect on the density near the Bondi radius.

The mass of the atmosphere typically determined by integrating the density profile from the core to the Bondi radius.  Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.  In this case
\begin{equation} \label{eq:MatmISO}
M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
\end{equation} 
This result is the leading order term in a series expansion.  Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.

From \Eq{eq:MatmISO} the ratio of isothermal atmosphere to core mass is 
\begin{equation} 
{M_{\rm iso} \over M\co} = 3 {\rho\di \over \rho\co}{R\co \over R_B} e^{R_B/R\co} %= {\theta_o \over \theta\co^2} e^{\theta\co}
\end{equation} 
While  $\rho\di \ll \rho\co$, the exponential dependence allows the atmosphere to become massive.  
%A value of $\theta\co = R_B / R\co$ corresponds to a planet mass of
%\begin{equation} 
%M\co = \sqrt{3 \over 4 \pi \rho\co} \left(\theta\co \mathcal{R} T \over G \right)^{3/2} 
%\end{equation} 
As a concrete example, consider a disk with $T = 60$ K and $\rho\di = 3 \times 10^{-11}$ (in cgs) and a core density $\rho\co = 2$.  The atmosphere self-gravitating for $\theta\co = 27.1$ or $M\co > 0.046 M_\oplus$.  The fact that very low mass cores can have very massive isothermal atmospheres is well known \citep{Sas89}.  However such solutions are unrealistic because there is inadequate time to radiate away all the atmosphere's thermal energy.

We not consider the mass exterior to the Bondi radius.  For a meaningful evaluation we only include the mass coming from the overdensity relative to the background density.  The resulting external mass for an isothermal atmosphere is
\begin{subeqnarray}
M_{\rm ext} &=& 4 \pi \int_{\RB}^{r_{\rm fit}} (\rho - \rho\di) r^2 dr \\
&=& M\co \theta\co \int_1^{n_{\rm fit}} 3 \left[ \exp \left({1 \over x} - {1 \over n_{\rm fit}}\right) - 1 \right] x^2 dx  \nonumber \\
&\equiv& M\co \theta\co I(n_{\rm fit})
\end{subeqnarray} 
where the dimensionless integral $I(n_{\rm fit})$ obeys the limits $I(1) = 0$ and $I \rightarrow n_{\rm fit}^2/2$ as $n_{\rm fit} \rightarrow \infty$.  Since this external mass does not converge the choice of an outer boundary does matter in principle.  In practice however the assumption that   $r_{\rm fit} = R_{\rm H}$ limits $n_{\rm fit}$ to modest values
\begin{equation}
n_{\rm fit} = {R_{\rm H} \over \RB} \approx 1.3 {\aun{10}^{4/7} \over \mcn{10}^{2/3}}{F_T \over  m_\ast^{1/3}}
\end{equation} 
since for instance $I(2) = 1.1$ these modest $n_{\rm fit}$ values will only produce a small external mass.

However the effect is still worth including for the effect on the interior mass.  
\subsection{Temperature Contrast at Convective Boundary}
We now show that the temperature contrast between the convective boundary, $T\cb$, and the ambient disk $T\di$, is modest.  We express the radiative lapse rate
\begin{equation}\label{eq:delrad}
\delrad \equiv {d \ln T \over d \ln P} = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P\di)^{1+\alpha} \over (T/T\di)^{4-\beta}}
\end{equation}
where $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.  If the mass in the radiative zone is small, then we can hold $M = M\co + M_{\rm conv}$ fixed at the sum of core and convective zone masses.  With this assumption and powerlaw opacity, we get a constant value for $\nabla\di$.  The temperature profile then integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T\di\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P\di}\right)^{1-\alpha} - 1 \right] \, ,
\end{equation} 
where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is $\delrad$ for $T ,P \rightarrow \infty$.\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to R06, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
We  apply \Eqs{eq:delrad}{eq:radTP} at the convective boundary $\delrad = \delad$ under the assumption that the pressure there is $P\cb \gg P\di$. The resulting temperature contrast at the convective boundary is
\begin{equation}\label{eq:Tcb}
\chi \equiv {T\cb\over T\di} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
\end{equation} 
For  the low temperatures in protoplanetary disks, opacity is dominated by dust with $\alpha = 0$ and $\beta \approx 2$ or $1$, giving $\chi \approx 1.5$ or 1.9, respectively, assuming  $\delad = 2/7$.

\subsection{Location of Convective Boundary}
The pressure at the convective boundary  follows from \Eqs{eq:delrad}{eq:Tcb} as
\begin{equation}
{P\cb\over P\di} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
\end{equation} 
This pressure contrast can be quite large due to the smallness of $\nabla\di$ in low luminosity atmospheres.

 It is also useful to obtain a relation between $T$ and $P$ that eliminates $\nabla\di$ in favor of $P\cb$:
 \begin{equation}\label{eq:TP}
{T \over T\di} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1-\alpha} -  \left({P\di \over P\cb}\right)^{1-\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
\end{equation}

We can determine the radius of the convective boundary $r\cb$ from the hydrostatic balance equation as 
\begin{equation}\label{eq:RCBint}
{R_B \over r\cb} = \int_{P\di}^{P\cb} {T \over T\di} {dP \over P}\, .
\end{equation} 
An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However using \Eq{eq:TP} in the integral gives
\begin{equation}\label{eq:Rcb}
{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
\end{equation} 
with an extra correction term, $\theta < 1$.  In the $P\cb \gg P\di$ limit the correction term is a constant that depends on $\alpha$, $\beta$ and $\delad$.  The form of $\theta$ was chosen to allow us to express
\begin{equation}\label{eq:PcbRcb}
P\cb = \theta P\di e^{R_B/r\cb} \, .
\end{equation}   
A simple analytic expression for $\theta$ is not possible.

\begin{deluxetable}{cccccc}  % <--- column justification (center/left/right)
\gdef \numcols {6}
\tablecolumns{\numcols}
\tablecaption{Parameters Describing Structure of Radiative Zone.}
\tablehead{   \multicolumn{\numcols}{c} {$\gamma = 7/5$ ($\delad = 2/7$), $\alpha = 0$} }  
\startdata
 $\beta$   		 &1/2  	& 3/4 &1   		& 3/2  		& 2   \\
 $\nabla_\infty$ & 2/7 \tablenotemark{a}  	&  4/13	& 1/3 	& 2/5 	 	& 1/2 \\
 $\chi$ 		 & \nodata &  2.25245 &1.91293 	& 1.65054 	& 1.52753 \\
 $\theta $  		 &\nodata   & 0.145032	&0.285824   &0.456333   & 0.556069   \\
\enddata
\tablenotetext{a}{Since $\delad = \nabla_\infty$ there is no convective transition at depth for this case.}
\end{deluxetable}

The T-P profile can also be combined with the equation of hydrostatic balance to relate the radius and pressure of the convective boundary.  If the convective boundary is well inside the Bondi radius, i.e.\ $r\cb\ll R_B$ and $P\cb\gg P\di$ then
\begin{equation}\label{eq:PcbRcb}
P\cb = \theta P\di e^{R_B/r\cb} \, .
\end{equation}    
The order unity constant $\theta < 1$ accounts for the fact that the radiative layer is not perfectly isothermal, as $\theta$ would be exactly one in that case.  The value of theta depends on the equation of state and the pressure and temperature dependence of the opacity, i.e. the $\alpha (=0)$ and $\beta$ values.  We find $\theta(\beta = 2) \approx 0.556$ and $\theta(\beta = 1) = 0.286$.  Closed form expressions for these integrals are too cumbersome to be useful.

\section{Virial Theorem}
The virial theorem for self-gravitating atmospheres can be applied to protoplanetary atmospheres.    The virial theorem is derived as usual by integrating the equation of hydrostatic balance to give
\begin{eqnarray}\label{eq:virial}
E_G= -3 \int_{M\co}^{M} {P \over \rho}dm + 4 \pi \left( R^3 P\surf - r\co^3 P\co\right) 
\end{eqnarray} 
where the gravitational energy
\begin{equation}\label{eq:EG}
E_G = -\int_{M\co}^{M} {G m \over r} dm \, .
\end{equation} 
At the outer surface $m = M$, $r = R$ and other values are given a subscripted $M$.  This outer surface can be evaluated anywhere (e.g.\ Bondi radius, Hill radius or RCB) that hydrostatic balance holds.  The boundary terms include the effects of finite surface pressure and finite radius of the solid core.

 The integral on the RHS \Eq{eq:virial} is related to the internal energy.  For an ideal gas, $P/\rho = \Rg T = (\gamma -1) u$, where $u$ is the internal energy per mass, and $\gamma = C_P/C_V$.  For a polytrope with constant $\gamma$ we can express the virial theorem in terms of internal energy $E_i = \int u dm$ as
\begin{equation}
E_G = - \zeta E_i + 4 \pi \left(R^3 P\surf - r\co^3 P\co\right) 
\end{equation} 
where $\zeta \equiv 3 (\gamma -1)$.  In this case the total energy $W = E_G + E_i$ becomes
\begin{subeqnarray}
W &=& (1-\zeta)E_i + 4 \pi \left( R^3 P\surf  - r\co^3 P\co \right) \\
 &=& {\zeta - 1 \over \zeta} E_G + {4 \pi \over \zeta} \left( R^3 P\surf  - r\co^3 P\co \right)\, .
\end{subeqnarray} 
The surface pressure acts to unbind the atmosphere by increasing $W$.  While a realistic EOS is neither ideal nor polytropic, \Eq{eq:virial} is general.

%\section{Adiabatic Atmosphere}
%An adiabatic atmosphere with the entropy of the disk gas is only slightly more massive than the disk gas itself (inside the Bondi radius).

\section{Cooling Models}
An isolated sphere satisfies a simple global energy equation
\begin{equation}\label{eq:simplecool}
L = \Gamma - \dot{E}
\end{equation} 
where the luminosity is balanced by the rate of heat generation, $\Gamma$, and the rate at which total (gravitational plus thermal, at least) energy is lost.

For a protoplanetary atmosphere embedded in a gas disk, the cooling equation is more complicated: 
\begin{equation}\label{eq:EB}
L = L\co + \Gamma- \dot{E} + e_{\rm acc} \dot{M}  -  P\surf {\p V\surf \over \p t} \, .
\end{equation}
We explain these extra terms and then derive the result.   \Eq{eq:EB} assumes a well (but arbitrarily) defined surface at radius $R$ and mass $M$, which can evolve in time.  Examples include the Bondi radius or the radiative-convective boundary.  The luminosity $L$ represents the luminosity from that surface.  The luminosity from the solid core is given by $L\co$, and includes planetesimal accretion energy  and radioactive decay.  The core is assumed to have a fixed mass, $M\co$, and radius, $R\co$, to simplify the atmospheric calculation.  The $\Gamma$ and $\dot{E}$ integrals are identical to \Eq{eq:simplecool}, and include the atmosphere from the core to the top boundary.   Sources of direct heating that  contribute to $\Gamma$ include  atmospheric drag on sedimenting planetesimals and the dissipation of any atmospheric turbulence.

Mass accretion at the rate $\dot{M}$ brings specific energy $e_{\rm acc} = u - GM/R$, the sum of internal and gravitational energies.  The energy of accreted matter, $e_{\rm acc}$, is  zero at a modified Bondi radius, $R_{\rm out} = G M/u \approx  (\Rg/C_V) R_B$.  For $R < R_{\rm out}$ ($R > R_{\rm out}$, respectively) the accreted mass has negative (positive) energy and is (un)bound.
 The final term in \Eq{eq:EB} gives the pressure work done on a surface mass element, so the partial time derivative of  volume holds the mass fixed.  Our quasistatic models we neglect bulk kinetic energy.

\subsection{Derivation of Global Energy Equation}
We now derive the global energy equation (\ref{eq:EB}).  We follow the simpler example in \S4.3 of \citet{KipWei94}, adding the effects of finite core radius, surface pressure and mass accretion.  We assume that hydrostatic balance holds.  Integrating the local energy equation from core to surface gives:
\begin{subeqnarray}
L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, .
\end{subeqnarray} 
with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate.

In what follows, we must carefully distinguish between partial time derivatives, $\p / \p t$, (performed at fixed mass) and total time derivatives, denoted with overdots (which include the effect of mass accreted through the outer boundary). % or $d/dt$.  
For instance the evolution of surface radius, $R$, evolves as  
\begin{equation}\label{eq:Rdot}
 \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho\surf}
\end{equation} 
where $\p R/\p t$ gives the Lagrangian contraction of surface mass elements, and $\dot{M}$ denotes mass accretion rate through the upper boundary.  The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  Similarly the volume, $V = (4 \pi/3)r^3$ and pressure at the surface evolve as
\begin{subeqnarray}\label{eq:dot}
\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
 \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} \\
 &=&  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
\end{subeqnarray} 
For the purpose of this derivation we will hold the core mass fixed $\dot{M}\co = 0$ which further gives $\dot{P}\co = \p P\co / \p t$.

To derive the global energy equation we must move the (partial) time derivatives in \Eq{eq:DLc} outside their integrals.  The internal energy integral follows simply from  Leibniz's rule as
\begin{equation}\label{eq:udot}
\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{E}_i  -  \dot{M}u\surf\, .
\end{equation}

To evaluate the work integral, we derive a pair of expressions for the rate of change of gravitational energy.
The time derivative of \Eq{eq:virial}  gives
\begin{eqnarray}\label{eq:EGdot}
\dot{E}_G &=& 3  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \\
&& -  3{P\surf \over \rho\surf} \dot{M}+ 3 \dot{P}\surf V\surf -3 \dot{P}\co V\co  + 3  P\surf {\dot{ V}\surf} \, . \nonumber
%&&+ 4 \pi \left. \left(r^3 {\p P \over \p t}\right)\right|_{M\co}^M  + 3  P\di {\p V \over \p t}  - {G M \over R} \dot{M} \nonumber
\end{eqnarray} 
%where the volumes, $V\surf = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
The first integral in \Eq{eq:EGdot} is the one we want, but the next one must be eliminated.  The time derivative of \Eq{eq:EG} (times four) gives
\begin{subeqnarray}
 4 \dot{E}_G &=&  -4 {G M \dot{M} \over R} + 4 \int_{M\co}^M {G m \over r^2}{\p r \over \p t} dm\\ 
&=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \slabel{eq:4EGc} \\
&&+ 3 V\surf {\p P\surf \over \p t} -3 V\co {\p P\co \over \p t} \nonumber 
\end{subeqnarray} 
where \Eqs{eq:4EGb}{eq:4EGc} use hydrostatic balance  and integration by parts.

%To eliminate the time derivates of pressure, we take the time derivative of the hydrostatic balance equation for $\p^2 P / \p m\p t$ and integrate over $4\pi r^3 dm$ (as in the virial equation derivation) to get
%\begin{equation}\label{eq:dHBdt}
%3 \dot{P}\surf V\surf -3 \dot{P}\co V\co -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  = 4 \dot{E}_G + 4{G M \over R} \dot{M}  \, .
%\end{equation} 
%Combining \Eqs{eq:EGdot}{eq:dHBdt} gives 
%\begin{eqnarray}\label{eq:rhodot}
%\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=& - \dot{E}_G - {4 \over 3}{G M\over R} \dot{M} + {P\surf \over \rho\surf} \dot{M} -  P\surf \dot{V}\surf  \, , \nonumber \\
%&=&- \dot{E}_G - {4 \over 3}{G M\over R} \dot{M}  -  P\surf {\p V\surf \over \p t}  \, ,
%\end{eqnarray} 
%where the final step uses \Eq{eq:Rdot}.

Subtracting \Eqs{eq:udot}{eq:4EGc} and rearranging for the desired integral gives
\begin{eqnarray}\label{eq:PdVint}
\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=&  - \dot{E}_G - {G M \dot{M} \over R} - P\surf {\p \dot{V}\surf \over \p t} \,  
\end{eqnarray} 
with \Eq{eq:dot} used to combine terms.  Combining \Eqsss{eq:DLc}{eq:udot}{eq:PdVint}, we reproduce \Eq{eq:EB} with the accreted energy $e_{\rm acc} \equiv u\surf - GM/R$.

>>>>

@@

\usepackage{pdfsync}
\usepackage{subeqnarray}
\usepackage{natbib}
\usepackage{color}

@@

\newcommand{\ie}{i.e.\ }
\newcommand{\eg}{e.g.\ }
\newcommand{\p}{\partial}
\newcommand{\brak}[1]{\langle #1\rangle}

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
\newcommand{\au}{\; \varpi_{\rm AU}}
\newcommand{\AU}{\; {\rm AU}}
\def\K{\; {\rm K}}

@@

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

\definecolor{gray}{gray}{0.5}
\newcommand{\emgr}[1]{\emph{ \color{gray} #1}}

@@

@@

\shorttitle{Critical Core Mass at Wide Separations}
\shortauthors{Piso \& Youdin}

\title{Lower Limits on the Core Mass for Giant Planets}
%\title{The Critical Core Mass of Wide-Separation Giant Planets:  Lower Limits from Kelvin-Helmholtz Cooling}
\author{Ana-Maria Piso}
\affil{Harvard Smithsonian Center for Astrophysics}
\author{ Andrew N.\ Youdin}
\affil{JILA, University of Colorado}
%\begin{abstract}
%\end{abstract}

\section{Introduction}
\emgr{Background topics to mention include: core accretion history, giant planets at wide separations and theoretical challenges posed (GI probably to big, core accretion maybe to slow).}

\emgr{Describe goals and approach of this paper.}

This paper is organized as follows: \emgr{(section summaries)}.

\section{Quasistatic Two Layer Model}\label{sec:model}

\subsection{Model Assumptions}
\emgr{Describe basic assumptions, perhaps in list format. Will need to describe basic scales and idea that atmosphere is 1D and matches smoothly onto disk.  However we may want to separate these standard assumptions from our more unique ones that are part of the two layer model.}

\subsection{Structure Equations}
\emgr{Give in pressure coordinates.  Put EOS (polytrope and detailed) and opacity laws (noting that we only consider cool radiative zones with dust opacities).}

\subsection{Boundary Conditions: Core and Disk}

%\subsection{Convective Zone}
%\emgr{might not need a (sub)section}

\subsection{Radiative Zone Solutions}
\emgr{Give analytic solution valid for power law opacity and lack of self-gravity.}

\subsection{Matching at the Radiative-Convective Boundary (RCB)}
\emgr{Include figure of an example atmospheric profile.}

\subsection{The Gravothermal Catastrophe}
\emgr{Show and explain double valued solutions, including caveat that upper branch breaks several model assumptions.}

\section{Cooling History}
\emgr{start with basic idea of quasistatic contraction and $L = -\dot{E}$}
\subsection{Role of Finite Pressure}
 \emgr{Give more more general result, but defer derivation to \App{sec:virial}.}
\subsection{Estimating a Critical Core Mass}
\emgr{Describe how and why we use the time to gravothermal catastrophe as estimate of minimum core mass}

\section{Critical Core Mass}
\emgr{These are the take home results!}

\subsection{Role of Disk Temperature and Pressure}

\subsection{Core Mass vs. Disk Radius}

\subsection{Opacity Dependence}

\subsection{EOS Tables}
\emgr{compare polytrope vs. real EOS, this could go first or last.}

\section{Conclusions}
